---
title: "Group Project 1"
subtitle: "Biology 368/664 Bucknell University"
output: pdf_document
authors: Prof. Ken Field
date: 14 Sep 2024
---

```{r Load Libraries, include=FALSE}
# Load other packages here.
if (!require("datasauRus")) install.packages("datasauRus"); library(datasauRus)
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse)
```

This project will require you to develop a tutorial to teach Bucknell students how to use R for graphing and data analysis. 

## Introduction

Begin by introducing yourself to your group. 
Then discuss the biggest challenge that you have faced during the first three weeks of this course.
Determine if there are any common threads in these challenges and start to think about objectives for the tutorial.

## Target Audience

Discuss with your group the target audience for the tutorial. 
Examples could be a new student in Biology 364/664, a student in one of the new core Biology classes (201, 202, 203, or 204), a student in another 300-level Biology course (not 364), or a new student in one of the Bucknell research groups. 

Edit the README.md file in your group's repository to reflect your plan.

## Objectives

After deciding on the target audience for your tutorial determine 2 to 3 overall objectives for your tutorial (one per member of your group). 
These should be high-level objectives that are important skills for your target audience.
Check with Prof. Field to see if they are appropriate and then add them to your README.md file.

Identify at least 2 goals within each objective and add them to your README.md file.
These should be goals that someone who is working through the tutorial can self-asses.
For example, "to demostrate that you can test a hypothesis using a statistical model, the student should use a T test, linear model, or other test and interpret the p value appropriately."

## Grading

Each group will be expected to complete the following tasks to earn 85% of the points available for this assignment (21/25).

- Identify and obtain suitable dataset
- Use a Github repository and version control to collaborate on the project
  + Every member of the group should participate in editing the repo
- Spend 4-6 hours preparing, coding, and testing tutorial
  + Data exploration
  + Data visualization
  + Hypothesis testing
- Present tutorial in class
- Provide public archive suitable for sharing to students/faculty


```{r - Converting to factor}
Esoph_scancer$agegp<- as.factor(Esoph_scancer$agegp)
Esoph_scancer$alcgp<- as.factor(Esoph_scancer$alcgp)
Esoph_scancer$tobgp<- as.factor(Esoph_scancer$tobgp)
```

```{r}
(esoph)
write.csv(esoph, "Esoph.scancer.csv", row.names = FALSE)
read.csv("Esoph.scancer.csv")
#testing for normality in the data
library(ggplot2)
library(cowplot)
a <- ggplot(Esoph_scancer, aes(x=tobgp)) + geom_bar() + ggtitle("Tobacco Group")
b <- ggplot(Esoph_scancer, aes(x=alcgp)) + geom_bar() + ggtitle("Alcohol Group")
c <- ggplot(Esoph_scancer, aes(x=agegp)) + geom_bar() + ggtitle("Age Group")

# Display the plots in a grid
plot_grid(a, b, c, ncol = 3)
```


#Making a histogram of the orginal and log transformed data and making density plots of them
```{r} 
# Create bar plots for discrete variables
a <- ggplot(Esoph_scancer, aes(x=tobgp)) + geom_bar() + ggtitle("Tobacco Group - Bar Plot")
b <- ggplot(Esoph_scancer, aes(x=alcgp)) + geom_bar() + ggtitle("Alcohol Group - Bar Plot")

# Create density plots using continuous data (if needed)
# Assuming there is a continuous variable, for example, ncases
# If not, density plots may not be applicable to this dataset

# For demonstration, let's assume a continuous variable ncases exists
c <- ggplot(Esoph_scancer, aes(x=ncases)) + geom_density() + ggtitle("Density Plot - ncases")
d <- ggplot(Esoph_scancer, aes(x=log(ncases))) + geom_density() + ggtitle("Density Plot - log(ncases)")

# Plot in a 2x2 grid
plot_grid(a, b, c, d, ncol = 2, nrow = 2)
```

```{r}
a<-qplot(data = Esoph_scancer,
         x=alcgp,
         geom = "histogram",
         bins=8)
b<-qplot(data = Esoph_scancer,
         x=log(alcgp),
         geom = "histogram",
         bins=8)
c<-qplot(data = Esoph_scancer,
         x=alcgp,
         geom = "density")
d<-qplot(data = Esoph_scancer,
         x=log(alcgp),
         geom = "density")
#Plot in a 2X2 grid

plot_grid(a,b,c,d, ncol = 2, nrow = 2)
```

#Normality Test: Shapiro test and fit distrubtions 
```{r}
shapiro.test(Esoph_scancer$alcgp)
shapiro.test(Esoph_scancer$tobgp)
```

```{r}
shapiro.test(log(Esoph_scancer$alcgp))
shapiro.test(log(Esoph_scancer$tobgp))
```

#finding error distributions

```{r}
library(MASS)
fit1<-fitdistr(Esoph_scancer$alcgp,"normal")
fit2<-fitdistr(Esoph_scancer$alcgp,"lognormal")
AIC(fit1,fit2)
```

###Data Visualization
Data visualization is another important aspect of data analysis because it allows you to simplify the data in a way that is easily digestible but still keep the importance of the findings. Another important reason to visualize the data is that most times, you will be presenting your findings to other people. There's also a good chance that your audience isn't as familiar with looking at data like you are so visuals are the best way to showcase your work.



